---
title: "Final Project"
author: "Evelyn Delph"
date: "4/21/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# library
library(tidyverse)
library(dplyr)
library(viridis)
library(broom)

# read in data + edits
df <- read.csv("master.csv")
df <- filter(df,gpaAverage >= 2)
df <- df %>%
  mutate(
    Class = recode_factor(Class,"Associate Professor" = "Assoc. Prof", "Assistant Professor" = "Asst. Prof")
  )
```

## Introduction

### Scope
* Grade distribution data from 2010-2018
* Professor salary data from 2010-2018
* Using gpaAverages between 2.0-4.0
* There are 13,065 rows of data

## Part 1: Exploratory Data Analysis
* First graph: Can we see a visible trend between gpa and salary over the 8 year span?
* Second graph: Can we see a visible trend between gpa, salary, and number of students per class type?
* Third graph: 


```{r,echo=F}
g1 <- df %>%
  select(c(Salary,gpaAverage,Year,Class)) %>%
  group_by(Year,Class) %>%
  summarise(aveGPA=mean(gpaAverage),aveSalary=mean(Salary)) %>%
  ggplot(aes(x=Year,y=aveGPA,color=aveSalary)) +
  geom_smooth(method="lm",se=FALSE,color="black")  +
  ggtitle("Increase in Salary and GPA Per Class Type") +
  xlab("Year")  + 
  ylab("Average GPA") + 
  geom_point(size=2,alpha=.4) + 
  facet_grid(~Class)  +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(color = "Avg Salary") + 
  scale_color_viridis_c() 

g2 <- df %>%
  group_by(Class,Year) %>%
  summarise(aveGPA=mean(gpaAverage),aveStudents=mean(nStudents),aveSalary=mean(Salary)) %>%
  ggplot(aes(x=aveGPA,y=aveStudents,color=aveSalary)) +
  geom_smooth(method="lm",se=FALSE,color="black")  +
  ggtitle("Relationship Between Average GPA & Number of Students Per Class Type") +
  xlab("Average GPA")  + 
  ylab("Average Number of Students") + 
  geom_point(size=2,alpha=.4) + 
  facet_grid(~Class)  +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(color = "Avg Salary") + 
  scale_color_viridis_c() 

g1
g2
```
```{r}
df %>%
  group_by(Class,Year,Org) %>%
  summarise(aveGPA=mean(gpaAverage),aveSalary=mean(Salary)) %>%
  ggplot(aes(x=aveGPA,y=aveSalary)) +
  geom_point() +
  facet_wrap(~Org)
```
Thomas...need help with codes
```{r,echo=F}
levels(unique(df$Org))
```

## Part 2: Models/Predictions


## Part 3: Analysis + Conclusion
```{r,echo=F}
```