---
title: "Final Project"
author: "Evelyn Delph, Thomas James, & Nick Mobley"
date: "4/21/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# library
library(tidyverse)
library(dplyr)
library(viridis)
library(broom)

# read in data + edits
df <- read.csv("master_file.csv")
depts <- read.csv("dept.csv")
df <- df %>%
  filter(gpaAverage >= 2) %>%
  filter(Course < 400) %>%
  mutate(
    Class = recode_factor(Class,"Associate Professor" = "Assoc. Prof", "Assistant Professor" = "Asst. Prof"),
    Level = round(Course,digits=-2)
  ) %>%
  filter(Level > 0) %>%
  inner_join(depts) %>% 
  select(c(Name,Category,Class,nStudents,Salary,Year,gpaAverage,Level))
```

# Introduction

## Abstract

Are undergraduate students receiving the upmost quality of education from there professors at Indiana University? Since instate students pay \$10,948 for tuition and out of state students pay \$36,512, it is important to determine if such expenditures are worth the cost. For the purposes of this report, we are examining if an instructors' grade distrubition for undergraduate courses is indicative instructors' salaries. For this study, the quality of education that an instructor provides is determined by their average GPA. We are interested in studying grade distrubtion and salary across other variables including: the department, the course level, the instructors' level of teaching, and the year. The goals of this report aim to answer the following questions:

* Does GPA impact an instructor's salary?
* Do GPA and salary relationships vary across course levels?
* Do GPA and salary relationships vary across years?
* Do GPA and salary relationships vary across different departments?

We will answer these questions with numerous EDA techniques to study the relationships among these variables. We aim to achieve a better understanding if the quality of education impacts instructors' salaries.

## Description of Data

Our master data file consists of 9 grade distribution and 9 salary files from the IU Registrar between the years 2010-2018. The datasets were joined by the instructor. The variables of interest pulled from these datasets include: Name of Instructor, Department, Instructor Level, Salary, Year, Average GPA, Number of Students, and Course Level.

```
DESCRIPTION
Total Rows: 15,163
Total Columns: 7
1. Name of Instructor: 1,942 unique instructors (string)
2. Departments: 100 departments at IU categorized into 8 groups (string)
3. Instructor Level: The title of an instructor (string)
4. Salary: The annual salary of a professor (integer)
5. Average GPA: Average GPA grouped by course and instructor, between 2.0-4.0 (double)
6. Number of Students: Number of students grouped by course and instructor (integer)
7. Course Level: Denotes 100, 200, 300, and 400 level classes (integer)
```

```{r,echo=F,fig.show = "hold",warning=F,out.width = "50%"}
ggplot(df,aes(x=log(Salary),fill=Class,color=Class)) +
  geom_density() + 
  facet_grid(~Class) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme(legend.position = "None",axis.text.x = element_text(angle = 45)) +
  ggtitle("Distribution of Salary Per Instructor Level")
  
ggplot(df,aes(x=gpaAverage,fill=Class)) + 
  stat_density() +
  scale_fill_viridis_d() +
  ggtitle("Distribution of Average GPA Per Instructor Level") +
  xlab("GPA") + 
  labs(fill = "Level")

qplot(df$Level,stat="count",fill=df$Class,binwidth=40) +
  ggtitle("Instructor Breakdown by Course Levels") +
  xlab("Course Level") + 
  ylab("counts") + 
  scale_fill_viridis_d() +
  labs(fill = "Level")

ggplot(df,aes(x=Class,fill=Category)) + 
 stat_count() +
 scale_fill_viridis_d() +
 ggtitle("Breakdown of Departments and Instructor Type") +
 labs(fill = "Departments") +
 xlab("Instructor Type")
```

# Modeling

...Have to work on this...

* First graph: Can we see a visible trend between gpa and salary over the 8 year span?
* Second graph: Can we see a visible trend between gpa, salary, and number of students per class type?

```{r,echo=F}
g1 <- ggplot(df,aes(x=gpaAverage,fill=Class)) + 
  stat_density() +
  scale_fill_viridis_d() +
  ggtitle("Distribution of Average GPA Per Class")

g1 <- df %>%
  group_by(Class) %>%
  ggplot(aes(x=gpaAverage,y=Salary,color=Class)) +
  geom_point(alpha=.1) + 
  geom_smooth(se=FALSE) +
  xlab("GPA") +
  ylab("Salary") +
  ggtitle("Average GPA and Average Salary Per Class Type")

g2 <- df %>%
  group_by(Class,Year) %>%
  ggplot(aes(x=gpaAverage,y=Salary,color=Class)) +
  geom_point(alpha=.1) + 
  geom_smooth(se=FALSE) +
  xlab("GPA") +
  ylab("Salary") +
  ggtitle("Average GPA and Average Salary Per Class Type") +
  facet_wrap(~Year)

g3 <- df %>%
  select(c(Salary,gpaAverage,Year,Class)) %>%
  group_by(Year,Class) %>%
  summarise(aveGPA=mean(gpaAverage),aveSalary=mean(Salary)) %>%
  ggplot(aes(x=Year,y=aveGPA,color=aveSalary)) +
  geom_smooth(method="lm",se=FALSE,color="black")  +
  ggtitle("Increase in Salary and GPA Per Class Type") +
  xlab("Year")  + 
  ylab("Average GPA") + 
  geom_point(size=2,alpha=.4) + 
  facet_grid(~Class)  +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(color = "Avg Salary") + 
  scale_color_viridis_c() 

g4 <- df %>%
  group_by(Class,Year) %>%
  summarise(aveGPA=mean(gpaAverage),aveStudents=mean(nStudents),aveSalary=mean(Salary)) %>%
  ggplot(aes(x=aveGPA,y=aveStudents,color=aveSalary)) +
  geom_smooth(method="lm",se=FALSE,color="black")  +
  ggtitle("Relationship Between Average GPA & Number of Students Per Class Type") +
  xlab("Average GPA")  + 
  ylab("Average Number of Students") + 
  geom_point(size=2,alpha=.4) + 
  facet_grid(~Class)  +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(color = "Avg Salary") + 
  scale_color_viridis_c()

g1
g2
g3
g4

```
```{r}

```

Thomas...need help with codes
```{r,echo=F}
levels(unique(df$Org))
```

## Part 2: Models/Predictions


## Part 3: Analysis + Conclusion
```{r,echo=F}
```